---
title: "Data Exploration"
author: "Ellen Bledsoe"
format: gfm
editor: visual
date: "`r Sys.Date()`"
---

## Data Exploration

First, load packages

```{r}
#| message: false
library(tidyverse)
```

### Read in Data Files

Explore the 4 CSV files from the Access Database.

```{r}
#| message: false
# List all CSV files in the folder
csv_files <- list.files("../data/", pattern = "\\.csv$", full.names = TRUE)

# Loop through each file
for (file in csv_files) {
  # Extract the filename without the .csv extension
  file_name <- tools::file_path_sans_ext(basename(file))
  
  # Read the CSV file and assign it to a variable with the extracted name
  assign(file_name, read_csv(file), envir = .GlobalEnv)
}
```

### Data Wrangling

First, get only data from *C. baileyi*. Combine tables to get survey information and dates with the records on individual rodents.

```{r}
# get only C. baileyi records
baileyi <- left_join(RodentDetail, tluRodentSpecies, 
                     join_by(RodentSpeciesID == ID)) |> 
  filter(SpeciesCode == "CHBA")

# get survey dates and sites with the baileys data
baileyi_surveys <- full_join(baileyi, RodentSurvey,
                             join_by(RodentSurveyID == ID)) |> 
  mutate(StartDate = ymd(StartDate),
         EndDate = ymd(EndDate)) |> 
  select(ID, RodentSurveyID, Recapture, SiteID:Quadrat)
baileyi_surveys
```

Summarize the survey data by year

```{r}
survey_by_year <- RodentSurvey |> 
  mutate(StartDate = ymd(StartDate),
         EndDate = ymd(EndDate),
         Year = year(EndDate),
         Nights = EndDate - StartDate)

trapnights_by_year <- survey_by_year |> 
  mutate(TrapNights = NumTraps * Nights) |> 
  group_by(Year) |> 
  summarise(TrapNights = as.numeric(sum(TrapNights))) 

trapnights_by_year
```

Determine the number of *C. baileyi* caught per year

```{r}
baileyi_year <- baileyi_surveys |> 
  mutate(Year = year(EndDate)) |> 
  group_by(Year) |> 
  summarise(Recaps = n(),
            NoRecaps = sum(Recapture == 0, na.rm = TRUE))

baileyi_year
```

Calculate *C. bailyei* per trapnight per year

```{r}
baileyi_trapnight <- inner_join(baileyi_year, trapnights_by_year,
                                join_by(Year)) |> 
  mutate(CHBA_TrapNight = Recaps / TrapNights,
         CHBA_TrapNight_NoRecaps = NoRecaps / TrapNights)

baileyi_trapnight
```

### Plot *C. baileyi* Numbers

Plots for number with and without including recaptures give similar results, so let's stick with no recaptures included

#### *C. baileyi* individuals per year

```{r}
ggplot(baileyi_trapnight, aes(Year, NoRecaps)) +
  geom_point() +
  geom_line() +
  theme_classic()
```

#### *C. baileyi* individuals per trap night

```{r}
ggplot(baileyi_trapnight, aes(Year, CHBA_TrapNight_NoRecaps)) +
  geom_point() +
  geom_line() +
  theme_classic()
```
